# Using ggplot with Other Packages

```{r setup, include = FALSE, purl = FALSE}
library(tidyverse)
library(ca)
library(ggrepel)
library(patchwork)
library(FactoMineR)
library(factoextra)
berry_data <- read_csv("data/clt-berry-data.csv")
raw_cider_data <- read_csv("data/CiderDryness_SensoryDATA.csv")
```

```{r data preprocessing}
#May replace some of this with read_rds() calls later
raw_cider_data %>%
  select(Sample_Name, Temperature, Fresh_Apples:Synthetic) %>%
  unite(Sample_Name, Temperature, col = "sample", sep = " ") %>%
  group_by(sample) %>%
  summarize(across(where(is.numeric), ~sum(.))) %>%
  column_to_rownames("sample") -> cider_contingency

cider_contingency %>%
  FactoMineR::CA(graph = FALSE) -> ca_cider

berry_data %>%
  pivot_longer(cols = matches("^(9pt|us|lms)_"),
               names_to = c("Scale", "Attribute"), names_sep = "_",
               values_to = "Liking",
               values_drop_na = TRUE) %>%
  select(where(~ !any(is.na(.x)))) %>%
  mutate(Liking = ifelse(Scale == "lms",
                         (Liking + 100) * 8/200 + 1,
                         ifelse(Scale == "us",
                                Liking * 8/15 + 1,
                                Liking))) %>%
  select(Subject = `Subject Code`, Sample = `Sample Name`, berry,
         starts_with("cata_"), Attribute, Scale, Liking) -> berry_data_long

berry_data_long %>%
  pivot_wider(names_from = Attribute, names_prefix = "liking_", values_from = Liking) %>%
  group_by(Sample) %>%
  summarize(across(starts_with("cata_"),
                   sum),
            across(starts_with("liking_"),
                   mean)) %>%
  select(where(~ !any(is.na(.x)))) %>%
  column_to_rownames("Sample") -> berry_mfa_summary

berry_mfa_summary %>%
  MFA(group = c(sum(str_detect(colnames(berry_mfa_summary), "^cata_")),
                sum(str_detect(colnames(berry_mfa_summary), "^liking_"))),
      type = c("f","s"), graph = FALSE,
      name.group = c("CATA","Liking")) -> berry_mfa_res
```

As you might have noticed, we had you download more packages than just `ggplot2` for this tutorial. `ggplot2` is a framework and will help you make many standard plots, but it can't do everything. Or, sometimes, you may not want to use it to do everything yourself.

Packages meant to work with `ggplot2` to more easily make specific kinds of visualizations are also called ggplot **extensions**. The four most common kinds of ggplot extensions are:

1. Packages that add `geom_*()s` or `stat_*()`s for new kinds of plots
2. Packages that add `theme_*()`s and `scale_*()`s for specific color or style needs
3. Packages that make `ggplot` objects, so you never write `ggplot()` yourself
4. Packages that combine multiple plots in various ways

You can view many of these extensions [on the tidyverse website](https://exts.ggplot2.tidyverse.org/gallery/) (where you'll also see many examples that fall into multiple of these categories or don't fit into the categories here at all). 

## New `geom_*()s` and `stat_*()`s
If you want to label each individual point in the plotting area using text, rather than some symbol or color that indicates the legend off to the side, you can do this using the base `ggplot2` functions `geom_text()` and `geom_label()`:

```{r using geom_text with 22 sensory attributes}
#Let's use the cider CA example from before.
#We can make our own plots from the coordinates.
ca_cider$col$coord %>%
  as_tibble(rownames = "Attribute") %>%
  ggplot(aes(x = `Dim 1`, y = `Dim 2`, label = Attribute)) +
  theme_bw() -> ca_cider_termplot

ca_cider_termplot +
  geom_text()
```

```{r using geom_label with 22 sensory attributes}
ca_cider_termplot +
  geom_label()
```
But, as you can see, the text starts to overlap itself quickly even with only a small handful of attributes. The extension I personally use most often, to make crowded plots like this more readable, is the package `ggrepel`, which adds new `geom_text_repel()` and `geom_label_repel()`.

```{r using geom_label_repel with 22 sensory attributes}
ca_cider_termplot +
  geom_label_repel()
```

They're almost identical to the normal text and label `geom_*()`s, but they use an iterative algorithm to push each piece of text away from the other text and unrelated points or `geom_()`s, while being pulled towards the point being labeled. It is *not deterministic*, so it will be slightly different each time you run the code (try it now!) unless you use `set.seed()` first or set `seed = ***` when adding `geom_*_repel()`.

Even with a set seed, changing the plot size or adding `geom_*()`s to the plot will also slightly change the locations of the labels, so if you're going to try and find a seed that works well for your data, you should be checking it on your final exported plot at the publication resolution (we'll talk about that next chapter).

```{r geom_text_repel with a set seed will look the same every time}
ca_cider_termplot +
  geom_label_repel(seed = 12345)
```

There are many settings you can play with to adjust these forces, how far a label has to move for a line to show up, whether any labels are left off in dense areas, and how long it tries to find a solution. And, randomly, the ability to give each letter-shape a border in a different color, which seems to be totally undocumented in the help files. It can be useful if there are other, multi-colored `geom_*()`s in the background.

```{r geom_text_repel text borders}
raw_cider_data %>%
  mutate(Product = str_c(Sample_Name, Temperature, sep = " ")) %>%
  group_by(Product) %>%
  summarize(Liking = mean(Liking)) %>%
  left_join(ca_cider$row$coord %>%
              as_tibble(rownames = "Product")) %>%
  rename_with(~ str_replace_all(.x, " ", ".")) -> ca_cider_productcoord

#This is NOT a statistically sound preference model, this is just for demonstration
ca_cider_prefmod <- lm(Liking ~ Dim.1 * Dim.2, data = ca_cider_productcoord)
expand.grid(Dim.1 = seq(min(ca_cider$col$coord[, "Dim 1"]) - 0.1,
                    max(ca_cider$col$coord[, "Dim 1"]) + 0.1,
                    by = 0.01),
            Dim.2 = seq(min(ca_cider$col$coord[, "Dim 2"]) - 0.1,
                    max(ca_cider$col$coord[, "Dim 2"]) + 0.1,
                    by = 0.01)) %>%
  mutate(., Liking = predict(ca_cider_prefmod, newdata = .)) -> ca_cider_prefinterp

ca_cider_termplot +
  geom_contour_filled(aes(x = Dim.1, y = Dim.2, z = Liking, fill = after_stat(level)),
                  inherit.aes = FALSE,
                  data = ca_cider_prefinterp) + 
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_text_repel(size = 6, color = "white", bg.color = "grey7")
```

Another useful tool for visualizing any ordinal/binned data (e.g., the 9-point hedonic scale) at scale is the `geom_beeswarm()` and `geom_quasirandom()` from the `ggbeeswarm` package, which are similar to `geom_jitter()` but intended for looking at a single numeric variable at a time, possibly across multiple categories.

They limit the jitter to a single direction and ensure that no points are overlapping (or, in the case of `geom_quasirandom()`, that there's a uniform amount of overlap) so you can get a more accurate picture of the density, but take up less space than many faceted `geom_histogram()`s (at least for the same amount of fine-tuning).

```{r using ggbeeswarm to compare ordinal distributions}
library(ggbeeswarm)
#The jitter plot is actually not very helpful with this many points
berry_data_long %>%
  ggplot(aes(x = Scale, y = Liking, color = Scale)) +
  geom_jitter() +
  facet_wrap(~ Attribute) +
  theme_bw()

#geom_beeswarm() will also have the same problem, but geom_quasirandom()
#visualizes the density at each "bin" without us having to specify bins.
#So these are easy to compare
berry_liking %>%
  ggplot(aes(x = Scale, y = Liking, color = Scale)) +
  geom_quasirandom() +
  facet_wrap(~ Attribute) +
  theme_bw()
```

If you want to easily add circles and ellipses to a plot (say, the unit circle or a confidence ellipse), you'll probably want to install the `ggforce` package and use `geom_circle()` or `geom_ellipse()`, respectively.

```{r adding a unit circle with ggforce}
berry_mfa_res$quanti.var$coord %>%
  as_tibble(rownames = "Modality") %>%
  ggplot() +
  geom_segment(aes(x = 0, y = 0, xend = Dim.1, yend = Dim.2), arrow = arrow()) +
  ggforce::geom_circle(aes(x0 = 0, y0 = 0, r = 1), color = "blue") +
  geom_text_repel(aes(x = Dim.1, y = Dim.2, label = Modality)) +
  theme_bw() +
  theme(aspect.ratio = 1)
```

## New `theme_*()`s and `scale_*()`s
Most of the additional `geom_*()`s in `ggplot2` extensions involve some sort of calculation, so the confidence that you're using someone else's algorithm that's (hopefully!) been double-checked is a real benefit. You've already seen how to change the way your plots look with `theme()` one argument at a time, and how to set `scale_*_manual()` if you have the exact colors or color range that you want.

So there's nothing these prettying-up packages will do that you can't do yourself, but there are a huge number of `ggplot2` extensions that include some version of a no-gridline minimal theme for convenience. Such as:

```{r the minimal themes in cowplot}
berry_liking %>%
  ggplot(aes(x = Scale, y = Liking, color = Scale)) +
  geom_quasirandom() +
  facet_wrap(~ Attribute) +
  cowplot::theme_minimal_hgrid()
```

Packages that added `scale_*()`s used to be one of the most common kinds of ggplot extensions (because, as you'll notice, the above figure with the default color scale is not red-green colorblind friendly), but the most popular scales now come with `ggplot2` itself.

`RColorBrewer`'s colorblind- and printer-friendly palettes for categorical data are now available in `ggplot2::scale_*_brewer_*()`, and you've already seen us use the `viridis` color palettes in `ggplot2::scale_*_viridis_*()`. The `viridis` color scales *can* be used for categorical data, if you use the `_d()` versions, but they were designed for ordinal and binned data, since some colors will seem more related than others. See [Chapter 4](https://clauswilke.com/dataviz/color-basics.html) and [Chapter 19](https://clauswilke.com/dataviz/color-pitfalls.html) of Claus O. Wilke's book *Fundamentals of Data Visualization*.

## Modifying `ggplot()`s made by other packages
So far in this chapter, we've been making all of the plots with a call to `ggplot()` and then adding on geoms, themes, labels, scales, and facets with `+`. But we've also been able to save our plots to variables partway through and then keep adding things to the saved plots. This is an incredibly useful difference from the way plots work in base R.

Some packages utilize `ggplot2` by making a whole plot for you with their own internal call to `ggplot()`, which means that you can treat it like any other ggplot for the sake of customizing. Most packages which can save a whole plot to a variable or output several plots in a list use `ggplot()` to do so.

```{r FactoMineR makes ggplots}
#FactoMineR uses ggplot for its internal plotting,
#Which is why we can assign the output to a variable
#and not see the plot right away
#(although the CA() function will also display several plots by default)
cider_contingency %>%
  FactoMineR::CA(graph = FALSE) %>%
  FactoMineR::plot.CA() -> ca_cider_biplot_facto

#The ca package, meanwhile, uses base plotting.
#You can tell because it prints this plot immediately.
cider_contingency %>%
  ca::ca() %>%
  ca::plot.ca() -> ca_cider_biplot_green
```

You can see that the last code chunk only output one plot right away, but we can confirm our suspicions with the base R `class()` function.

```{r using class to find ggplots}
class(ca_cider_termplot) # Made with ggplot() ourselves
class(ca_cider_biplot_facto) # Made with ggplot-based FactoMineR
class(ca_cider_biplot_green) # Made with ca::ca(), not a plot at all
ca_cider_biplot_green # It's two tables of coordinates!
```

What this means is that we can look at the `FactoMineR`-made plot we've saved to `ca_cider_biplot_facto`:

```{r you'll see this ggplot now}
ca_cider_biplot_facto
```

*And* we can still change up many of the elements by adding additional elements, although you're likely to get some weird warning messages *and* some silent errors.

```{r you can change many things about FactoMineR plots with ggplot semantics}
ca_cider_biplot_facto +
  theme(panel.grid = element_blank(), # Removes the axis lines
        plot.title = element_blank()) + # Removes the title
  xlim(-1,1) + # Extends the x limits, with a warning
  scale_color_brewer(type = "qual", palette = "Dark2")
  # Silently fails to change the color scale
```

You can't just `+ scale_color_*()` to a `FactoMineR` plot, because the `ggplot` already has a non-default color scheme and adding a second color scale does nothing. If you look at the help file for `?FactoMineR::plot.CA`, you can set several styling parameters when you're making the plot, and you can remake it as many times as you need, but doing so does have significantly less flexibility than the approach to plotting we've outlined in this workshop.

We also can't go back and adjust the parameters passed to `geom_text_repel()` after the fact, even though we can tell from the warning messages that that's the package being used to put the attribute and product names onto the biplot.

It will almost always be possible to add more `geom_*()`s to plots made by other packages, as long as you don't mind them being added *on top of* any existing elements in the plot.

```{r you can easily add new geoms to FactoMineR plots}
liking_arrow <- data.frame(x1 = 0, y1 = 0, x2 = -0.4, y2 = -0.1, text = "Liking")

ca_cider_biplot_facto +
  geom_segment(aes(x= x1, y = y1, xend = x2, yend = y2), color = "orange",
               arrow = arrow(length = unit(0.03, "npc")), linewidth = 1,
               data = liking_arrow) +
  geom_text(aes(x = x2, y = y2, label = text), color = "orange",
            hjust = "outward", vjust = "outward", fontface = "italic",
            data = liking_arrow)
```

If you desperately need to [change a scale or reorder `geom_*()`s from an existing `ggplot`](https://cran.r-project.org/web/packages/gginnards/vignettes/user-guide-2.html#replacing-scales-coordinates-whole-themes-and-data.) in a hurry, look into the `gginnards` package.

## Combining Plots
You've already seen how to `facet_*()` plots to view "small multiple" plots side-by side:

```{r remember faceting, fig.align = "center"}
raw_cider_data %>%
  pivot_longer(Fresh_Apples:Synthetic) %>%
  group_by(Sample_Name, Temperature, name) %>%
  summarize(total = sum(value)) %>%
  ggplot(aes(x = interaction(Sample_Name, Temperature), y = total)) + 
  geom_col(aes(fill = Sample_Name)) +
  scale_fill_manual(values = wesanderson::wes_palettes$FantasticFox1) + 
  coord_flip() + 
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_bw() + 
  theme(legend.position = "top",
        panel.grid = element_blank()) -> cider_count_plot

cider_count_plot +
  facet_wrap(~name, ncol = 6)
```

This works very well whenever you have multiple plots using the same `geom_*()`s that you want to show on the same axes, and you can even adjust the axis limits from facet to facet using `scales = "free*"`:

```{r faceted plots with different x-axes}
cider_count_plot +
  facet_wrap(~name, ncol = 6,
             scales = "free_x") # Each plot now has a different x-axis
```

Not that we'd argue you *should* here. Also, take note that the `x` in `free_x` refers to the horizontal axis in the final plot, *after* the `coord_flip()`, and *not* the `x` aesthetic we set in the `ggplot()` call.

But if you have different plot types entirely (different data sources, different `geom_()`s, or different categorical axes) that you want to place side-by-side, say a loading plot and the product map resulting from a PCA or MFA, you're going to need something to paste together multiple `ggplot_()`s.

The easiest way to do this is using `patchwork`, which will work on ggplots you've made yourself *or* with ones made by packages like `FactoMineR`. *When you have `patchwork` loaded*, the `+` operator will put two plots side-by-side:

```{r combining plots with patchwork}
library(patchwork)
plot(berry_mfa_res, choix = "var") + plot(berry_mfa_res, partial = "all")
```

And the `/` operator will arrange two plots vertically:

```{r arranging plots vertically with patchwork}
plot(berry_mfa_res, choix = "var") / plot(berry_mfa_res, partial = "all")
```

The advantage of doing this with a package like `patchwork`, rather than saving separate images, is that it aligns all of the plot areas precisely and that they will more easily move or rearrange certain plot elements like legends and axis labels.

```{r collecting legends at the bottom of a patchwork ensemble}
plot(berry_mfa_res, choix = "var") + plot(berry_mfa_res, partial = "all") +
  plot_layout(guides = "collect") &
  theme(plot.title = element_blank(),
        legend.position = "bottom")
```

The `&` operator lets you add elements like themes or annotations to all of the plots you've combined together. `plot_layout()` is a `patchwork` function that lets you set relative plot sizes, decide how to arrange more than 2 plots, and move legends:

```{r more complex patchwork layout}
plot(berry_mfa_res, partial = "all") +
  (plot(berry_mfa_res, choix = "var") +
  plot(berry_mfa_res, choix = "freq", invisible = "ind")) +
  plot_layout(guides = "collect", ncol = 1, widths = 2) &
  theme(plot.title = element_blank(),
        axis.title = element_blank(),
        legend.position = "bottom")
```

If you have an R `list` structure containing multiple plots, you want to put images anywhere on a visualization, or you're struggling to make a complex arrangement with `patchwork`, `cowplot` is another option:

```{r }

```

Both of these packages can also add letters and other labels to each plot:

```{r }
plot(berry_mfa_res, choix = "var") + plot(berry_mfa_res, partial = "all") +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A') &
  theme(plot.title = element_blank(),
        legend.position = "bottom")


```

If you need to move or realign the labels so they're not overlapping anything, you can use <> in `patchwork` and <> in `cowplot`:

```{r }

```

You can also use any image editing, publishing, or graphics software to manually combine, arrange, and label plots, but if you need to make changes to a plot later then doing your layout *in R* will mean you just have to run the lightly-updated code again to re-export a fully formatted multi-part figure, even if the plot dimensions change.

## Finding New Packages

This isn't all the functionality of any of these packages, and these aren't the only packages that add new features to `ggplot2`.

If you're trying to figure out how to create a type of plot you've never made before, we'd recommend:

1. Ask yourself what variables are represented by the x and y axes, the shapes, the line types, or the colors. Think about whether you can build the plot from smaller components. Is it a grid of scatterplots with colored contour regions? How much of the plot is just points, lines, or rectangles with fancy formatting? You can do **a lot** with just `ggplot2`!
2. See if any of the packages in [the list of registered ggplot extensions](https://exts.ggplot2.tidyverse.org/gallery/) have a plot similar to yours. These packages tend to have very thorough and visual documentation.
3. If it's a sensory-specific plot, [check out the R Opus v2](https://jlahne.github.io/r-opus-v2/index.html). Almost all of the plots only use `ggplot2`, `patchwork`, and `ggrepel`.
4. Use a web search with the kind of plot you want to make and the keyword "ggplot2" to find tutorials or discussions with example code. Results from [Stack Overflow](https://stackoverflow.com/questions/tagged/ggplot2), [Data to Viz](https://www.data-to-viz.com/), or [R Graph Gallery](https://r-graph-gallery.com/index.html) are all likely to have good explanations and useful examples, while anything with "(# examples)" in the article name is likely to be very basic material with good SEO.

Keep in mind that keywords primarily used in sensory science, say "preference map" or "penalty analysis", are unlikely to yield examples in `ggplot2` with extensive results. In my own process of troubleshooting and double-checking for this workshop, I've found some helpful examples by searching "add density contours to 2d scatterplot ggplot" or "ordered bar plot positive and negative ggplot".

Package documentation might come up while you're looking, but the examples are often very abstract and simple, and they're often structured around names of functions rather than concepts, so it's often faster to see some examples of real plots that other people (who aren't writing an entire R package) wanted to make and then looking up the functions they used to do so.

Now, with the rest of the time we have left, I'm going to try and show off how I do some of the most common cleaning-up when it's time to actually present or publish.