# Fine-Tuning Publication-Quality ggplots

```{r setup, include = FALSE, purl = FALSE}
library(tidyverse)
library(ca)
library(ggrepel)
library(patchwork)
library(FactoMineR)
library(factoextra)
berry_data <- read_csv("data/clt-berry-data.csv")
```

The plots we've been making so far are fairly straightforward demonstrations. So far, the most complicated plots we've all made together looks like this:

```{r penalty analysis example}
ggplot(mapping = aes(x = cata_variable, y = penalty_lift)) +
  geom_col(aes(fill = penalty_lift), color = "white", show.legend = FALSE) + 
  facet_wrap(~berry, scales = "free", nrow = 1) + 
  tidytext::scale_x_reordered() + 
  coord_flip() + 
  theme_classic() + 
  scale_fill_gradient(low = "tan", high = "darkgreen") + 
  labs(x = NULL, y = NULL,
       title = "Penalty / Lift Analysis",
       subtitle = "displays the mean difference (within berries) for when a CATA variable is checked\nor un-checked")
```



In an actual publication, where space is precious and we're trying to show complex data relationships, we might be looking at something more like this:

```{r final graphs}

```

## Exporting and Saving Plots

It may seem like a weird time to be talking about saving plots, when we haven't gotten to the "fine-tuning" yet, but you should already be familiar with a few examples of things that <>. It is basically impossible to export a `ggplot` at the resolution needed for publishing without changing the aspect ratio, relative sizes, or space between some elements from what you're looking at in the Plots pane or the `.Rmd` chunk output.

It's good practice to export your plot as an image (or however else you plan to export it) and re-export it periodically as you go through fine-tuning.

This is also, probably, the most important part of this chapter. You will have to get every plot that you publish out of `R` somehow, after all! The rest of the advice in this chapter are tricks that we draw upon frequently, but you may find that you have different problems or priorities, or that some of these needs are situational.

But you will have to export plots for every project you want to share with the world.

## Making Text Look Okay

### Powerful Text Manipulation with `stringr`

### Special characters, Spaces, and Line Breaks

### Adjusting Text Alignment

## Adding, Removing, and Styling Legends and Plots with `theme()`

## Ordered Categorical Variables

### Specifying Ordinal Variables as Factors

### Alternative Categorical-Ordering Options

## Combining Many `geom_*()`s or `aes()`thetics

