# Reading and Looking at Data in R

```{r setup, include = FALSE, purl = FALSE}
library(tidyverse)
library(ca)
berry_data <- read_csv("data/clt-berry-data.csv")
```

## The "working directory" and why you should care

Before we move on to using `R` for real, we have one key general computing concept to tackle: the "working directory".  The working directory is the folder on your computer in which `R` will look for files and save files.  When you need to tell `R` to read in data from a file or output a file, you will have to do so **in relation to your working directory**.  Therefore, it is important that you know how to find your working directory and change it.

The easiest (but not best) way to do this is to use the `Files` pane.  If you hit the "gear" icon in the `Files` pane menu, you'll see two commands to do with the working directory.  You can `Go To Working Directory` to show you whatever `R` currently has set as the working directory.  You can then navigate to any directory you want on your hard drive, and use the `Set As Working Directory` command to make that the working directory.

A better way to do this is to use the R commands `getwd()` and `setwd()`.

```{r}
getwd() # will print the current working directory
```

And we can manually change the working directory by using 

```{r, eval=FALSE}
setwd("Enter/Your/Desired/Directory/Here")
```

Notice that I am not running the second command, because it would cause an error!

When we use `R` to navigate directories, I recommend **always** using the forward slash: `/`, even though on Windows systems the typical slash is the backslash: `\`.  `R` will properly interpret the `/` for you in the context of your operating system, and this is more consistent with most modern code environments.

## Reading data into `R`

So far we've only done very basic things with `R`, which probably haven't sold you on its power and utility.  Let's try doing some things that will hopefully get us a little further towards actual, useful applications.  

First off, make sure you have the `tidyverse` package loaded by using the `library()` function.

```{r loading packages 1}
library(tidyverse)
```

Now, we're going to read in the data we're going to use for this workshop using the function `read_csv()`.  If you want to learn about this function, use the `?read_csv` command to get some details.

In the workshop archive you downloaded, the `data/` directory has a file called `clt-berry-data.csv`.  If you double-click this in your file browser, it will (most likely) open in Excel, and you can take a look at it.  The data describe the attributes and liking scores reported by consumers for a variety of berries across multiple CLTs. A total of 969 participants (`Subject Code`) and 23 berries (`Sample Name`) were involved in these tests, with only one species of berry (blackberry, blueberry, raspberry, or strawberry) presented during each CLT.  In the actual experimental design, subjects got multiple sample sets (so there are *not* 969 unique subjects), but here we will treat them as unique for ease of description.

As for the survey, panelists were asked JAR, CATA, and free response questions, as well as liking on the Unlabeled Line Scale (`us_*`), 9-point scale (`9pt_*`), or labeled magnitude scale (`lms_*`).

To get this data into `R`, we have to tell `R` where it lives on your computer, and what kind of data it is.

### Where the data lives

We touched on **working directories** because this is how `R` "sees" your computer.  It will look first in the working directory, and then you will have to tell it where the file is *relative* to that directory.  If you have been following along and opened up the `.Rproj` file in the downloaded archive, your working directory should be the archive's top level, which will mean that we only need to point `R` towards the `data/` folder and then the `clt-berry-data.csv` file.  We can check the working directory with the `getwd()` function.

```{r getting the working directory}
getwd()
```

Therefore, **relative to the working directory**, the file path to this data is `data/clt-berry-data.csv`.  Please note that this is the UNIX convention for file paths: in Windows, the backslash `\` is used to separate directories.  Happily, RStudio will translate between the two conventions, so you can just follow along with the macOS/UNIX convention in this workshop.

### What kind of file are we importing?

The first step is to notice this is a `.csv` file, which stands for **c**omma-**s**eparated **v**alue.  This means our data, in raw format, looks something like this:

```
# Comma-separated data

cat_acquisition_order,name,weight\n
1,Nick,9\n
2,Margot,10\n
3,Little Guy,13\n
```

Each line represents a row of data, and each field is separated by a comma (`,`).  We can read this kind of data into `R` by using the `read_csv()` function.

```{r reading in data}
read_csv(file = "data/clt-berry-data.csv")
```

Suddenly, we have tabular data (i.e., data in rows and columns), like we'd have in Excel!  Now we're getting somewhere.  However, before we go forward we'll have to store this data somewhere--right now we're just reading it and throwing it away.

```{r storing data when loaded}
berry_data <- read_csv(file = "data/clt-berry-data.csv")
```

As a note, in many countries the separator (delimiter) will be the semi-colon (`;`), since the comma is used as the decimal marker.  To read files formatted this way, you can use the `read_csv2()` function.  If you encounter tab-separated values files (`.tsv`) you can use the `read_tsv()` function.  If you have more non-standard delimiters, you can use the `read_delim()` function, which will allow you to specify your own delimiter characters.  You can also read many other formats of tabular data using the `rio` package ("read input/output"), which can be installed from CRAN (using, as you have learned, `install.packages("rio")`).

## Looking at Data

`berry_data` is a type of object called a `tibble`, which is a cute/fancy version of the more basic `R` object called a `data.frame`.  These are `R`'s version of the `.csv` file or your typical Excel file: a rectangular matrix of data, with (usually) columns representing some variable and rows representing some kind of observation.  Each row will have a value in each column or will be `NA`, which is `R`'s specific value to represent **missing data**.

In a `data.frame`, every column has to have only a single data type: a column might be `logical` or `integer` or `character`, but it cannot be a mix.  However, each column can be a different type.  For example, the first column in our `beer_data`, called `reviews`, is a `character` vector, but the third column, `beer_id`, is a `numeric` column.  

We have now moved from 1-dimensional vectors to 2-dimensional data tables, which means we're going to have some new properties to investigate.  First off, we might want to know how many rows and columns our data table has:

```{r data frame properties}
nrow(berry_data)
ncol(berry_data)
length(berry_data) # Note that length() of a data.frame is the same as ncol()
```

We already tried running `str(berry_data)`, which gives us the data types of each column and an example.  Some other ways to examine the data include the following:

```{r examining what a data frame contains, eval = FALSE}
berry_data # simply printing the object
head(berry_data) # show the first few rows
tail(berry_data) # show the last few rows
glimpse(berry_data) # a more compact version of str()
names(berry_data) # get the variable/column names
```

Some of these functions (for example, `glimpse()`) come from the `tidyverse` package, so if you are having trouble running a command, first make sure you have run `library(tidyverse)`.

## Subsetting and wrangling data tables

Since we now have 2 dimensions, our old strategy of using a single number to select a value from a vector won't work!  But the `[]` operator still works on data frames and tibbles.  We just have to specify coordinates, as `[<row>, <column>]`.  

```{r subsetting data tables 1}
berry_data[3, 9] # get the 3rd row, 9th column value
```

We can continue to use ranges or vectors of indices to select larger parts of the table

```{r subsetting data tables 2}
berry_data[1:6, 9] # get the first 5 rows of the 9th column value
```

If we only want to subset on a specific dimension and get everything from the other dimension, we just leave it blank.

```{r subsetting data tables 3}
berry_data[, 1] # get all rows of the 1st column
```

We can also use logical subsetting, just like in vectors.  This is very powerful but a bit complicated, so we are going to introduce some `tidyverse` based operators to do this that will make it a lot easier.  I will just give an example:

```{r logical subsetting for data tables}
berry_data[berry_data$berry == "raspberry", ] # get all raspberry data
```

In this last example I also introduced the final bit of `tibble` and `data.frame` wrangling we will cover here: the `$` operator.  This is the operator to select a single column from a `data.frame` or `tibble`.  It gives you back the vector that makes up that column:

```{r getting variables out of a data frame, eval = FALSE}
berry_data$lms_overall # not printed because it is too long!
berry_data$`9pt_overall` # column names starting with numbers or containing special characters like spaces need to be surrounded by backticks (`)
```

One of the nice things about RStudio is that it provides tab-completion for `$`.  Go to the console, type "ber" and hit `tab` (or just wait a second or two, if you use the default RStudio settings).  You'll see a list of possible matches, with `berry_data` at the top.  Hit enter, and this will fill out the typing for you!  Now, type "$" and hit `tab` again.  You'll see a list of the columns in `berry_data`!  This can save a huge amount of typing and memorizing the names of variables and objects.

Now that we've gone over the basics of creating and manipulating objects in `R`, we're going to run through the basics of data manipulation and visualization with the `tidyverse`.  Before we move on to that topic, let's address **any questions**.